An efficient composition of bidirectional programs by tupling and lazy updates

### 2020-02-18 paper/2.minbigul.tex
explain examples + revise experiments descr.

\noindent As an example of minBiGUL program, consider the definition of $phead$ :\\
\smallvspace
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$
\smallvspace

@@

Next let us look at another more complex example using a case combinator:

    $\tab bsnoc = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
    $\tabs{3} \ cond_{sv} = \lambda s.\lambda v.(\textnormal{length } v = 1), \ cond_s = \lambda (s_1,s_2).(s_2 = [\ ])$\\
    $\tabs{3} \ bx_1 =  Replace$\\
    $\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
        $\tabs{4} f_1 = f_2^{-1} = \lambda (x:y:ys).(y,(x:ys))$, \\
        $\tabs{4} bx_s = RearrV \ g_1 \ g_2 \ bx_v$ where:\\
            $\tabs{5} g_1 = g_2^{-1} = \lambda (v:vs).(v,vs)$\\
            $\tabs{5} bx_v = Replace \times bsnoc$

To close this section,
% In the end of this section, 
let us see the minBiGUL definition of $\mathit{bfoldr}$ which is a putback function of an important higher-order function on lists, $\mathit{foldr}$:

\smallvspace
$\tab \mathit{bfoldr} \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times bfoldr \ bx) \circ bx)$
\smallvspace

\noindent In $\mathit{bfoldr}$, the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. Because~$\circ$ has a higher priority than~$\times$, it is in general not possible to transform $\mathit{bfoldr}$ from the left associative composition style to the right one. 
%will look like a composition of many programs at some points in the evaluation if we slow down this process. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one.

>>>>

\noindent As an example of minBiGUL program, consider the definition of $\textit{phead}$ :\\
\smallvspace
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$
\smallvspace

@@

In the same way with $\textit{phead}$, we can define $\textit{ptail}$ in minBiGUL. $\putbx{ptail}{s}{v}$ accepts a source list $s$ and a view list $v$ to produce a new list by replacing the tail of $s$ with $v$. $\getbx{ptail}{s}$ returns the tail of the source list $s$

Next let us look at another more complex example, $\textit{bsnoc}$:

    $\tab \textit{bsnoc} = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
    $\tabs{3} \ cond_{sv} = \lambda s.\lambda v.(\textnormal{length } v = 1), \ cond_s = \lambda s.(\textnormal{length } s = 1)$\\
    $\tabs{3} \ bx_1 =  Replace, \tab bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
        $\tabs{4} f_1 = f_2^{-1} = \lambda (x:y:ys).(y,(x:ys))$, \\
        $\tabs{4} bx_s = RearrV \ g_1 \ g_2 \ bx_v$ where:\\
            $\tabs{5} g_1 = g_2^{-1} = \lambda (v:vs).(v,vs), \tab bx_v = Replace \times \textit{bsnoc}$

$\putbxinline{bsnoc}$ requires the source $s$ and the view $v$ are non-empty lists and the length of $v$ is not larger than the length of $s$. If $cond_{sv}$ is true, i.e. $v$ is singleton, a replacement will be executed to produce a new list which should be equal to $v$. Because the length of the new list is 1, the exit condition $cond_s$ comes true, so we obtain the updated source. If $v$ is a list of more than one elements, there will be two rearrangements on the source and the view before conducting a product. The program rearranges the source $x:y:ys$ to a pair of its second element $y$ and a list $x:ys$ created from the remaining elements in the original order, and the view to a pair of its head and tail. Then we can use $y$ to replace the head of the view and pair $(x:ys)$ with the tail of the view to form the input of a recursive call $\textit{bsnoc}$. The obtained source update in this case should be non-singleton since the value of the exit condition $cond_s$ needs to be false. We omit the behavior description of $\getbxinline{bsnoc}$ that accepts a source list $s$, checks $cond_s$ to know how to evaluate the view $v$, then does one more checking, $cond_{sv}$, before resulting. 
%
Intuitively, $\putbx{bsnoc}{s_0}{v_0}$ produces a new list by moving the last element of $v_0$ to its first position if the length of $v_0$ is not larger than the length of $s_0$. $\getbx{bsnoc}{s_0}$ returns another list by moving the first element of the list $s_0$ to its end position. For instance, $\putbx{bsnoc}{[1,2,3]}{[4,5,6]} = [6,4,5]$ and $\getbx{bsnoc}{[1,2,3]} = [2,3,1]$.

Now, let us see the minBiGUL definition of $\textit{bfoldr}$ which is a putback function of an important higher-order function on lists, $\textit{foldr}$:

$\tab \textit{bfoldr} \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times bfoldr \ bx) \circ bx)$

If we think that a minBiGUL program $bx$ has a type of $\textit{MinBiGUL} \ s \ v$, the type of $bfoldr$ will be look like $\textit{MinBiGUL} \ (a, b) \ a \to \textit{MinBiGUL} \ ([a], b) \ b$. You can easily find the similarity between the above definition of $\textit{bfoldr}$ with the following definition of $\textit{foldr}$:

$\tab foldr :: (a \to b \to b) \to b \to [a] \to b \\
\tabs{2} \ \ foldr \ f \ e \ [\ ] = e \\ 
\tabs{2} \ \ foldr \ f \ e \ (x:xs) = f \ x \ (foldr \ f \ e \ xs)$

Each branch in a case of $\textit{bfoldr}$ corresponds to a pattern of $foldr$. In $\textit{bfoldr}$, the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. Because~$\circ$ has a higher priority than~$\times$, it is in general not possible to transform $\textit{bfoldr}$ from the left associative composition style to the right one.

Using $\textit{foldr}$, we can define other functions, for example:

$\tab reverse = foldr \ snoc \ [\ ]$

With $\textit{bfoldr}$, we can also write the bidirectional version $breverse$ as follows:

$\tab breverse = RearrS \ f_1 \ f_2 \ bx$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda s \to (s, [\ ]), \tab bx = bfoldr \ bsnoc$

### 2020-02-18 paper/4.xpg.tex
tiny + update review-todo
explain examples + revise experiments descr.

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{3} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tabs{2} \qtab (s_3, v_3') \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{3} (s_3, v_2)$

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
%We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
Since there is no real view, this $dummy$ is necessary to pair with original source $s$ to form the input of $put \, [\![bx_1]\!]$. The last equality changes two. The first is a change from $dummy$ to an application $construct\_dummy \ s$. This means we need a way to construct $dummy$ from the current source, $s$. Programmers can be required to give this $construct\_dummy$ function, but it may be inessential for ill-typed systems where choosing $dummy$ as $v$ is the easiest way to meet our expectation. We set up that choice in our experiments but still use the $construct\_dummy$ function in the later definitions to provide a more general meanning.
%
The second is the change from $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ to $(s_3, v_3') \Leftarrow \pg{bx_1}{s_1}{s_2}$, where $s$ and $v_3$ are replaced with $s_1$ and $v_3'$ respectively. Because $s_1$ is a source update of $\putbx{bx_1}{s}{dummy}$, so under the \textsc{PutPut} law, it is possible to substitute $s$ by $s_1$ is possible. The substitution of $v_3$ by $v_3'$ is simply replacing the variable name since $v_3$ and $v_3'$ hold different results of $\getbx{bx_1}{s}$ and $\getbx{bx_1}{s_1}$ respectively. Because both variables are no longer used later, this substitution does not affect the outcome of the function.

@@

    \noindent $\cpg{bx_1 \circ bx_2}{ks}{kv}{s}{v} =\\
        \tab (ks_1, kv_1, \underline{s_1}, v_1) \Leftarrow \cpg{bx_1}{ks}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{kv_1}{kv}{v_1}{v};\\
            \qtab (ks_1 \circ ks_2, kv_2,  ks_1 \ s_2, v_2)$
\end{definition}

In the places where third and/or fourth argument ($s$ and $v$) are updated by applications, the computations are also accumulated in $ks$ and/or $kv$.
%Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.
Thanks to these accumulations, there are only two $cpg$ calls in $\cpginline{bx_1 \circ bx_2}$. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source same as $pg$. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

@@

    \noindent $\kpg{bx_1 \circ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab (ks_1, kv_1, \underline{ks_1'}, kv_1', \underline{s_1}, v_1) \Leftarrow \kpg{bx_1}{ks}{id}{ks'}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, ks_2', kv_2', s_2, v_2) \Leftarrow \kpg{bx_2}{kv_1}{kv}{kv_1'}{kv'}{v_1}{v};\\
        \qtab (ks_1 \circ ks_2, kv_2, ks_1 \circ ks_2', kv_2', s_2, v_2)$
\end{definition}

@@

    \noindent $\xpg{bx}{s}{v} = \match{bx}\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } pg$
\end{definition}

>>>>

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{3} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ v}; \\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tabs{2} \qtab (s_3, v_3') \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{3} (s_3, v_2)$

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
%We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
Since there is no real view, this $dummy$ is necessary to pair with original source $s$ to form the input of $put \, [\![bx_1]\!]$. The last equality changes two. The first is a change from $dummy$ to an application $construct\_dummy \ v$. This means we need a way to construct $dummy$ from the original view $v$. Programmers can be required to give this $construct\_dummy$ function, but it may be inessential for ill-typed systems where choosing $dummy$ as $v$ is the easiest way to meet our expectation. We set up that choice in our experiments but still use the $construct\_dummy$ function in the later definitions to provide a more general meanning.
%
The second is the change from $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ to $(s_3, v_3') \Leftarrow \pg{bx_1}{s_1}{s_2}$, where $s$ and $v_3$ are replaced with $s_1$ and $v_3'$ respectively. Because $s_1$ is a source update of $\putbx{bx_1}{s}{dummy}$, so under the \textsc{PutPut} law, it is possible to substitute $s$ by $s_1$. The substitution of $v_3$ by $v_3'$ is simply replacing the variable name since $v_3$ and $v_3'$ hold different results of $\getbx{bx_1}{s}$ and $\getbx{bx_1}{s_1}$ respectively. Because both variables are no longer used later, this substitution does not affect the outcome of the function.

@@

    \noindent $\cpg{bx_1 \circ bx_2}{ks}{kv}{s}{v} =\\
        \tab (ks_1, kv_1, \underline{s_1}, v_1) \Leftarrow \cpg{bx_1}{ks}{id}{s}{construct\_dummy \ v};\\
        \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{kv_1}{kv}{v_1}{v};\\
            \qtab (ks_1 \circ ks_2, kv_2,  ks_1 \ s_2, v_2)$
\end{definition}

In the places where third and/or fourth argument ($s$ and $v$) are updated by applications, the computations are also accumulated in $ks$ and/or $kv$.
%Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.
Thanks to these accumulations, there are only two $cpg$ calls in $\cpginline{bx_1 \circ bx_2}$. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ v)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source same as $pg$. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

@@

    \noindent $\kpg{bx_1 \circ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab (ks_1, kv_1, \underline{ks_1'}, kv_1', \underline{s_1}, v_1) \Leftarrow \kpg{bx_1}{ks}{id}{ks'}{id}{s}{construct\_dummy \ v};\\
        \tab (ks_2, kv_2, ks_2', kv_2', s_2, v_2) \Leftarrow \kpg{bx_2}{kv_1}{kv}{kv_1'}{kv'}{v_1}{v};\\
        \qtab (ks_1 \circ ks_2, kv_2, ks_1 \circ ks_2', kv_2', s_2, v_2)$
\end{definition}

@@

    \noindent $\xpg{bx}{s}{v} = \match{bx}\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ v};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } pg$
\end{definition}

### 2020-02-18 paper/5.experiments.tex
explain examples + revise experiments descr.

$\putbx{ptail}{s}{v}$ accepts a source list $s$ and a view list $v$ to produce a new list by replacing the tail of $s$ with $v$. $\getbx{ptail}{s}$ returns the tail of the source list $s$. The composition of many $ptail$s, in the \textit{put} direction, replaces a part of the tail of the source list by the view list and, in the \textit{get} direction, returns such a tail from the source.

$\putbx{bsnoc}{s}{v}$ accepts a source $s$ and a view $v$, 
%as two same length lists, 
which are two lists of the same length,
and produces a new list by moving the last element of $v$ to its first position. $\getbx{bsnoc}{s}$ returns another list by moving the first element of the list $s$ to its end position. The composition of many \textit{bsnoc}s, in the \textit{put} direction, creates a permutation of the view list if the source list has a same length as the view and, in the \textit{get} direction, produces a permutation of the source list.

\textit{breverse} is defined in terms of \textit{bfoldr}. In the \textit{put} direction, it produces a reverse of the view list if the source list has a same length as the view and, in the \textit{get} direction, produces a reverse of the source list. Note that compositions are by the recursions of $breverse$ and 
the number of compositions are dynamically determined by the length of the source list.

>>>>

The composition of many $ptail$s, in the \textit{put} direction, replaces a part of the tail of the source list by the view list and, in the \textit{get} direction, returns such a tail from the source.

The composition of many \textit{bsnoc}s, in the \textit{put} direction, creates a permutation of the view list if its length is not larger than the length of the source list and, in the \textit{get} direction, produces a permutation of the source list.

\textit{breverse} is defined in terms of \textit{bfoldr}. In the \textit{put} direction, it produces a reverse of the view list if its length is not larger than the length of the source list and, in the \textit{get} direction, produces a reverse of the source list. Note that compositions are by the recursions of \textit{breverse} and 
the number of compositions are dynamically determined by the length of the source list.
